---
title: "N.C. Police Certification Initial Data Exploration"
author: "Lucas Thomae"
date: "`r Sys.Date()`"
output: html_document
---

The [Wandering Officer article](https://www.yalelawjournal.org/article/the-wandering-officer) published in the Yale Law Journal by Ben Grunwald and John Rappaport in April 2020 defines wandering officers as "law-enforcement officers fired by one department, sometimes for serious misconduct, who then find work at another agency."

Earlier this year, we obtained a dataset from the North Carolina Department of Justice that tracks actions taken on N.C. police officer's certifications going back to 1973, when this data collection started. That data frame is loaded in as `nc_cert`.

It is our goal to identify wandering officers in North Carolina.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(stringdist)
library(DT)

nc_cert <- readxl::read_xlsx("~/Documents/Fall 2023/Ind. Study/NC_Police_Cert_Request_Data.xlsx")
officer_counts_2022 <- readxl::read_xlsx("~/Documents/Fall 2023/Ind. Study/2022 Law Enforcement Personnel By Agency.xlsx")
```

This is a check to see if `ACADEMYID` is a unique identifier for each officer in the data set.

```{r}
length(unique(nc_cert$ACADEMYID))
length(unique(nc_cert$NAME))
```

This checks out. `ACADEMYID` has 158 more unique values because of some repeat names. For my analysis purposes going forward, I will use `ACADEMYID` for cleaning the data.

Now, I'll make a dataframe containing only active officers. `nc_active` will contain active officers and all of the actions taken on their certifications. Then, I can count the number of unique values in the ACADEMYID column to count the number of individual officers in the data that had an active certification at the time of the data pull.

```{r}
nc_active <-
  nc_cert %>%
  group_by(ACADEMYID) %>% #grouping by individuals
  filter(any(grepl("General|Probationary|Grandfathered", CURRENTCERTSTATUS, ignore.case = F))) %>%
  ungroup()

length(unique(nc_active$ACADEMYID))
```

There are 30,931 unique ACADEMYID values in `nc_active`. We've also obtained data, contained in the data frame `officer_counts_2022`, which lists the number of sworn officers at each law enforcement agency in North Carolina as of 2022. We can compare those numbers to the one in the `nc_active` data frame as a check to see how reliable the data in `nc_cert` is.

```{r}
officer_counts_2022 <-
  officer_counts_2022 %>%
  mutate(
    "Sworn Officers 2022" = `Fulltime Male Sworn` + `Fulltime Female Sworn`
  ) %>%
  select("Agency", "Sworn Officers 2022")
```

```{r}
sum(officer_counts_2022$`Sworn Officers 2022`)
```

There were 24,180 active law enforcement officers in North Carolina in 2022, according to `officer_counts_2022`. We know that law enforcement agencies across the country have shrank year to year since 2020 and struggled to fill vacancies, so we couldn't expect that number to jump up to 30,931 in 2023. It appears that at least some of the data in `nc_cert` is incomplete or inaccurate, which has caused us to over-estimate the active officers in North Carolina by about 6,000.

I'll continue to run checks on the data, this time counting the number of officers in each agency.

```{r}
department_counts <-
nc_active %>%
  group_by(CERTIFICATION) %>%
  filter(grepl("(Active)", CURRENTCERTSTATUS, ignore.case = F)) %>%
  distinct(ACADEMYID, CERTIFICATION, .keep_all = TRUE) %>%
  summarise(officers = n()) %>%
  arrange(desc(officers))

department_counts
```

On a glance over, this looks solid to me. As I'd expect, Charlotte-MEcklenburg PD has the most officers, followed by the State Highway Patrol and the Raleigh PD. I'll further compare these numbers to the ones in `officer_count_2022` by joining the two data frames by the `CERTIFICATION` and `Agency` columns. The values in those two columns represent the same agencies, but the strings themselves are styled differently (For example, "New Hanover County Sheriff's Office - Deputy Sheriff" and "New Hanover Co. Sheriff"), so I'll have to use the `stringdist` package to match the string values up as best I can and then join the data frames.

```{r}
threshold <- 0.4

matches <- stringdist::amatch(department_counts$CERTIFICATION, officer_counts_2022$Agency, method = "jw", maxDist = threshold)

# Create a new dataframe with matched agency names
matched_df <- data.frame(Department = department_counts$CERTIFICATION,
                         Agency = officer_counts_2022$Agency[matches])

# Perform a left join with officer_counts_2022
joined_df <- left_join(department_counts, matched_df, by = c("CERTIFICATION" = "Department"))

joined_df <- left_join(joined_df, officer_counts_2022, by = "Agency") %>%
  rename(
    "Estimated Officers" = "officers",
    "Matched Agency" = "Agency",
  ) %>%
  mutate("Difference" = `Estimated Officers` - `Sworn Officers 2022`,
         "% Diff" = (`Difference` / `Sworn Officers 2022`)*100)

datatable(joined_df)
```

This join gives me a clearer picture of what's going on. While the estimates for some agencies like Charlotte-Meck PD and the NCSHP are very close to each other, the estimates for some agencies like the Buncombe County Sheriff's Office way overshot the true officer count.

My guess is that this is a result of some agencies not consistently or accurately reporting when an officer separates from the department, and therefore they are left in the database and appear to be an active officer. I am awaiting clarification from Russell Henderson, the research specialist at the NCDOJ for an explanation of the disagreement between these numbers.

This data may be messy, but I still want to examine individual officers in the data to see of any anomalies pop out. A starting hypothesis of mine is that officers with many actions in the data and who have been with many different departments deserve a more scrutinous eye when looking for 'wandering officers'.

I'll use `group_by()` and `summarize()` to count the number of actions taken on each officer's certification and also count the number of departments they've been with. Since certifications in North Carolina are tied to the specific agency that officers are employed by, we can use the CERTIFCATION column to count the departments.

```{r}
nc_active %>%
  group_by(ACADEMYID) %>%
  summarize(actions = n(),
            departments = length(unique(CERTIFICATION))) %>%
  arrange(desc(actions))
```

The above dataframe captures all 30,931 active officers in the cleaned data. To narrow that down, I'll filter `nc_active` to only include active officers who at one point have had an inactive certification at some point in their employment history. That would include officers who've had their `ACTIONSTATUS` marked as "Denied", "Separated", "Pending Inactive" or "Suspended".

```{r}
currentactive_inactives <- 
nc_active %>%
  group_by(ACADEMYID) %>%
  filter(any(grepl("(Inactive)", ACTIONSTATUS, ignore.case = F)))

currentactive_inactives %>%
  group_by(ACADEMYID) %>%
  summarize(actions = n(),
            departments = length(unique(CERTIFICATION))) %>%
  arrange(desc(actions))
  
```

```{r}
nc_active %>%
  filter(ACADEMYID == "0030-1622") %>%
  arrange(ACTIONEFFECTIVEDATE) %>%
  select(NAME, CERTIFICATION, ACTIONSTATUS, ACTIONEFFECTIVEDATE, CURRENTCERTSTATUS) %>%
  datatable()
```

Now I'll repeat this process, but filter by specific `ACTIONSTATUS` values like "Suspended" and "Revoked".

```{r}
currentactive_suspended <- 
nc_active %>%
  group_by(ACADEMYID) %>%
  filter(any(grepl("Suspended", ACTIONSTATUS, ignore.case = F)))

currentactive_suspended %>%
  group_by(ACADEMYID) %>%
  summarize(actions = n(),
            departments = length(unique(CERTIFICATION))) %>%
  arrange(desc(actions))
```

```{r}
nc_active %>%
  filter(ACADEMYID == "0036-7316") %>%
  arrange(ACTIONEFFECTIVEDATE) %>%
  select(NAME, CERTIFICATION, ACTIONSTATUS, ACTIONEFFECTIVEDATE, CURRENTCERTSTATUS) %>%
  datatable()
```

I'll explore another active officer who's had their certification revoked in the past.

```{r}
currentactive_revoked <- 
nc_active %>%
  group_by(ACADEMYID) %>%
  filter(any(grepl("Revoked", ACTIONSTATUS, ignore.case = F)))

currentactive_revoked %>%
  group_by(ACADEMYID) %>%
  summarize(actions = n(),
            departments = length(unique(CERTIFICATION))) %>%
  arrange(desc(actions))
```

```{r}
nc_active %>%
  filter(ACADEMYID == "0040-2559") %>%
  arrange(ACTIONEFFECTIVEDATE) %>%
  select(NAME, CERTIFICATION, ACTIONSTATUS, ACTIONEFFECTIVEDATE, CURRENTCERTSTATUS) %>%
  datatable()
```

